CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-07-10 script/en/chapters/numbertheory.tex
no message

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}
{chapter \ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix D of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
we have\footnote{%
  See \hyperlink{AppArith2}{Appendix D of this chapter} for source code implementing the
  square and multiply method in Mathematica and Pari-GP, which can be used
  to reproduce the calculations above. }:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}

@@

{\bf Example 1:} \\
The following table\footnote{%
See \hyperlink{AppArith3a}{Appendix D of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$ and for the bases 
$a =  1, 2, \cdots, 10$ as well as the resulting value $ord_{11}(a)$ for each $a$:

@@

{\bf Example 2:} \\
The following table\footnote{%
See \hyperlink{AppArith3b}{Appendix D of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and for the bases 
$a =  1, 2, \cdots, 12$ as well as the resulting value $ord_{45}(a)$ for each $a$:

@@

{\bf Example 4:} \\
The following table\footnote{%
See \hyperlink{AppArith3c}{Appendix D of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} answers the question as to whether the number $7$ is a primitive root of $46$.
The requirement/condition $gcd(7,46)=1$ is fulfilled.

@@

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3,713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix D of this chapter} for source code to do RSA
  encryption using Mathematica and Pari-GP.  You can also encrypt
  the message with CrypTool\index{CrypTool} v1.3 {\bf Indiv. Procedures \textbackslash{} RSA Demonstration \textbackslash{} Factorisation of a Number}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503, q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix D of this chapter} for the source code to factorise
the number $J(n)$ using Mathematica and Pari-GP.
Using CrypTool\index{CrypTool} v1.3 you can solve this with the 
{\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} 
Factorisation of a Number}.}. \\
{\bf 2.} $e=256,027$ (should lie between $509$ and $255,016$ and must be relatively prime to $255,016$)\footnote{$e$ cannot,
therefore, be $2, 127$ or $251$ ($65537 = 2^{16}+1$).\\
In real life, $J(n)$ is not factorised but rather the Euclidean algorithm 
is used for the selected e to guarantee that ${\rm gcd}(d,J(n))=1$.}.\\
{\bf 3.} $d=231,953$ \\\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}67,000)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

\label{SrcArith4c}
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix D of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

\label{SrcArith4d}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4d}{Appendix D of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt 158721 137346 37358 240130 112898}

@@

\label{SrcArith4e}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4e}{Appendix D of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt 198967 051405 254571 115318 014251}

@@

\newpage
\subsection*{Appendix D: Examples using Mathematica and Pari-GP} \hypertarget{AppendixD}{}
\addcontentsline{toc}{subsection}{Appendix D: Examples using Mathematica and Pari-GP}
\index{Mathematica}\index{Pari-GP}

>>>>

Cryptographically not only the unique nature of the inverse is important, but
also that the set of possible values has been exhausted.
\begin{theorem}\label{thm-zth-exhperm}
For $a,i\in \{1, \cdots, m-1 \}$ with ${\rm~gcd} (a,m)=1),$ then the product $a*i {\rm ~mod ~} m $ takes for a 
certain number $a$ all values from $\{1, \cdots ,m-1 \} $ (exhaustive permutation \index{Permutation} of the length $m-1$)\footnote{See
also theorem \ref{thm-zth-ordp} in \hyperlink{Kap_3_9}
{chapter \ref{MultOrdPrimitveRoot}, Multiplicative order and primitive roots}.}.
\end{theorem}
\vskip +10 pt
The following three examples\footnote{%
See \hyperlink{AppArith1}{Appendix E of this chapter} for the source code to compute
the tables using Mathematica\index{Mathematica} and Pari-GP.\index{Pari-GP}
}
illustrate the properties of multiplicative inverses.

@@

and the squares (mod $103$) can be calculated beforehand
\begin{eqnarray*}
87^2 & \equiv & 50 {\rm ~(mod~}103),\\
87^4 & \equiv & 50^2 \equiv 28 {\rm ~(mod~}103), \\
87^8 & \equiv & 28^2 \equiv 63 {\rm ~(mod~}103), \\
87^{16} & \equiv & 63^2 \equiv 55 {\rm ~(mod~}103),\\
87^{32} & \equiv & 55^2 \equiv 38 {\rm ~(mod~}103).
\end{eqnarray*}
we have\footnote{%
  See \hyperlink{AppArith2}{Appendix E of this chapter} for source code implementing the
  square and multiply method in Mathematica and Pari-GP, which can be used
  to reproduce the calculations above. }:
\label{SrcArith2}
\begin{eqnarray*}
87^{43} & \equiv & 87^{32+8+2+1}{\rm ~(mod~}103) \nonumber \\
        & \equiv & 87^{32} * 87^8 * 87^2 * 87 {\rm ~(mod~}103) \nonumber \\ 
    & \equiv & 38 * 63 * 50 * 87 \equiv 85 {\rm ~(mod~}103). \nonumber
\end{eqnarray*}

@@

{\bf Example 1:} \\
The following table\footnote{%
See \hyperlink{AppArith3a}{Appendix E of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} 
shows the values $a^i$ mod $11$ for the exponents $i = 1, 2, \cdots, 10$ and for the bases 
$a =  1, 2, \cdots, 10$ as well as the resulting value $ord_{11}(a)$ for each $a$:

@@

{\bf Example 2:} \\
The following table\footnote{%
See \hyperlink{AppArith3b}{Appendix E of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} shows the values $a^i$ mod $45$ for the exponents $i = 1, 2, \cdots, 12$ and for the bases 
$a =  1, 2, \cdots, 12$ as well as the resulting value $ord_{45}(a)$ for each $a$:

@@

{\bf Example 4:} \\
The following table\footnote{%
See \hyperlink{AppArith3c}{Appendix E of this chapter} for the source code to generate
the table using Mathematica and Pari-GP.
} answers the question as to whether the number $7$ is a primitive root of $46$.
The requirement/condition $gcd(7,46)=1$ is fulfilled.

@@

\label{SrcArith4a}
All 7 parts are encrypted using: $C \equiv M^{37}{\rm ~(mod~}3,713)$\footnote{%
See \hyperlink{AppArith4a}{Appendix E of this chapter} for source code to do RSA
  encryption using Mathematica and Pari-GP.  You can also encrypt
  the message with CrypTool\index{CrypTool} v1.3 {\bf Indiv. Procedures \textbackslash{} RSA Demonstration \textbackslash{} Factorisation of a Number}.}: \\
{\tt 1404 2932 3536 0001 3284 2280 2235}

@@

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503, q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix E of this chapter} for the source code to factorise
the number $J(n)$ using Mathematica and Pari-GP.
Using CrypTool\index{CrypTool} v1.3 you can solve this with the 
{\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} 
Factorisation of a Number}.}. \\
{\bf 2.} $e=256,027$ (should lie between $509$ and $255,016$ and must be relatively prime to $255,016$)\footnote{$e$ cannot,
therefore, be $2, 127$ or $251$ ($65537 = 2^{16}+1$).\\
In real life, $J(n)$ is not factorised but rather the Euclidean algorithm 
is used for the selected e to guarantee that ${\rm gcd}(d,J(n))=1$.}.\\
{\bf 3.} $d=231,953$ \\\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}67,000)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

@@

\label{SrcArith4c}
Each character is encrypted using: $C = M^{65,537} {\rm ~(mod~} 256,027)$\footnote{%
See \hyperlink{AppArith4c}{Appendix E of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt
\begin{tabular}{lllll}
212984 & 025546 & 104529 & 031692 & 248407 \\
100412 & 054196 & 100184 & 058179 & 227433\\
\end{tabular} }

@@

\label{SrcArith4d}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4d}{Appendix E of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt 158721 137346 37358 240130 112898}

@@

\label{SrcArith4e}
Each block is encrypted using: $C \equiv M^{65,537}{\rm ~(mod~}256,027)$\footnote{%
See \hyperlink{AppArith4e}{Appendix E of this chapter} for the source code for RSA 
exponentiation using Mathematica and Pari-GP.
}: \\
{\tt 198967 051405 254571 115318 014251}

@@

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\subsection*{Appendix D: Base representation of numbers}
\addcontentsline{toc}{subsection}{Appendix D: Base representation and base transformation of numbers}

For a given number $z$ one may ask how to represent such a number. In general we use
representations like $z = 2374$ or $z = \sqrt{2}$. The second number consists of an infinite
number of digits and therefore it can never be described precisely by the first representation. In this case 
the number must rounded.

We represent numbers usually in the decimal system (base 10). Computers
are working with the binary representation of numbers --- only for the display numbers are 
represented in decimal or sometimes hexadecimal (base 16).

This appendix describes how to generate arbitrary base representations of a positive integers and how 
to determine the number of required digits via the log function.

\subsubsection*{$b$-adic sum representation of positive integers}
Given base $b$, each positive integer $z$ can be represented as a $b$-adic sum 
$$ z = a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0, $$
where $a_i \in \{0, 1, \dots, b-1\}, \; i=0,1,\dots, n$ are called {\em digits}.
For this sum, it follows that for arbitrary digits $a_0, a_1, \dots, a_n$ we have 
$b^{n+1} > a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0$. On the other hand there exists
$a_0, a_1, \dots, a_n$ (namely $a_i = b-1$ for $i=0, \dots, n$) following that  
$b^{n+1} -1 \le a_nb^n + a_{n-1}b^{n-1} + \cdots + a_1b + a_0$
(using those inequalities it can be shown that each positve integer can be represented by a $b$-adic sum).

By writing the digits $a_na_{n-1} \cdots a_1a_0$ in a row the usual writing for numbers comes to hand.

{\bf Examples}: \\
Base $b = 10$: $10278 =  1\cdot10^4 +  0\cdot 10^3 + 2\cdot 10^2 + 7\cdot 10^1 + 8$ und \\
Base $b = 16$: $FE70A = 15\cdot16^4 + 14\cdot 16^3 + 7\cdot 16^2 + 0\cdot 16^1 + 10$.

\subsubsection*{Number of digits to represent a positive integer}

For a positive integer $z$ the length of the $b$-adic representation can be determined via the 
following steps. Starting from the inequality $b^{n+1} > z \ge b^n$ 
we have after applying the logarithm function on basis $b$ \footnote{%
Applying the logarithm formula on base $b$ and $b'$ we have $\log_b z = \log_{b'} z / \log_{b'} (b)$. 
It is therefore easy using e.g. logarithm tables for the base $b' = 10$ to compute the logarithm of base $b = 2$.
} $n+1 > log_b z \ge n$. Therefore we have $n = \lfloor log_b z \rfloor$. 
We call $l_b(z)$ {\em the number of required digits to represent the number $z$ on the base $b$}. We have
$$l_b(z) := \lfloor log_b z \rfloor +1.$$

\vskip +10 pt
{\bf Example}:
The decimal number $z = 234$ we compute for the hexadecimal representation (base of numbers $b = 16$)
$$l_{16}(z) = \lfloor \log_{16}(z) \rfloor + 1 = \lfloor \ln (z) / \ln(16) \rfloor + 1 = \lfloor 1,96... \rfloor + 1 = 1 + 1 = 2.$$

\subsubsection*{Algorithm to compute the base representation}
Given the number $z$ one can compute the base $b$ representation of $z$ using the following algorithm

\begin{tabbing}
bla\= \kill
{\bf input}: $z, b$ \\
$n := 0, z' := z$ \\
{\bf while} $z' > 0$ {\bf do} \\
 \> $a_n := z' {\rm ~mod~} b$,\\
 \> $z' := \lfloor z' / b \rfloor$  \\
 \> $n := n+1$ \\
{\bf end do} \\
{\bf output}:  $a_n a_{n-1} \cdots a_1 a_0$ in base $b$ representation. 
\end{tabbing}

{\bf Example 1}:
The integer $z = 234$ on the base of numbers $10$ will be transformed into the hex representation via
$a_0 = 234 {\rm ~mod~} 16 = 10 = A$, $234 / 16 = 15 = E$, $a_1 = 14 {\rm ~mod~} 16 = E$ and therefore we have $EA$.

\vskip +10 pt
{\bf Example 2}:
The binary number $z = 1000100101110101$ is computed into the decimal representation via the following steps:

$1000100101110101 = 1001$ (mod $1010)  \Longrightarrow a_0 = 9$, $1000100101110101 / 1010 = 110110111110$ \\
$110110111110 = 1000$ (mod $1010) \Longrightarrow a_1 = 8$, $110110111110 / 1010 = 101011111$ \\
$101011111 = 1$ (mod $1010) \Longrightarrow a_2 = 1$, $10101111 / 1010 = 100011$ \\
$100011 =  101$ (mod $1010) \Longrightarrow a_3 = 5$, $100011 / 1010 = 1$ \\
$11 = 11$ (mod $1010) \Longrightarrow a_4 = 3$ \\
therefore $z = 35189$.

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\subsection*{Appendix E: Examples using Mathematica and Pari-GP} \hypertarget{AppendixD}{}
\addcontentsline{toc}{subsection}{Appendix E: Examples using Mathematica and Pari-GP}
\index{Mathematica}\index{Pari-GP}
